---
title: "ARIMA Time Series Implementation"
output:
  word_document: default
  pdf_document: default
  html_notebook: default
---
This is an implementation of [ARIMA](https://www.rdocumentation.org/packages/forecast/versions/8.12/topics/Arima) model for forecasting future sales for a retailer. The data consists product sales information of united states. There are three product categories mentioned in the data set Office supplies, Technology and Furniture

```{r}
# Importing required libraries
library(tseries)
library(forecast)
library(dplyr)
library(xlsx)
library(readxl)
library(ggplot2)

# Assigning working directory
setwd("F:/Mayur/ANALYTICS/Data scientist/Data scienc with R/Time series")

# importing data
df = read_excel("Sample - Superstore.xls",sheet = "Orders")
View(df)
glimpse(df)

```

---
Checking missing values present in the data set
```{r}
colSums(is.na(df)) # no missing values
```

---
Exploring sales variable
```{r}
summary(df$Sales)

boxplot(df$Sales, main = "Product sales in United States", ylab="Sales in $",
        color = "orange", border = "brown",
        horizontal = FALSE, notch = TRUE)

hist(df$Sales,breaks = 20,col = "skyblue3",main = "Histogram of united stste sales",xlab = "Sales in $")

## observations with sales >=12000
View(df %>% filter(Sales>=10000)) # techonology related products

```

---
Checking unique product list
```{r}
length(unique(df$`Product Name`))
length(unique(df$`Product ID`)) 
# similar product id is assigned to different product names
```

---
Checking uniqueness of customer id and customer name
```{r}
length(unique(df$`Customer ID`))
length(unique(df$`Customer Name`))

```

---
Further exploration of product id and product list
```{r}
product_list = df %>% group_by(`Product ID`,`Product Name`) %>% 
            summarise(count = n()) %>% arrange(`Product ID`)

length(unique(product_list$`Product Name`))
length(unique(product_list$`Product ID`))
dim(product_list)

product_id = product_list %>% group_by(`Product ID`) %>% mutate(count = row_number())

## finding out the prducts with similar product id
similar_id = product_id %>% filter(count >= 2)

View(df %>% filter(`Product ID` %in% similar_id$`Product ID`
                  ) %>% 
         arrange(`Product ID`))

View(df %>% filter(`Product Name`%in% similar_id$`Product Name`) %>% 
    arrange(`Product ID`))

# keeping the records though it has similar product id
# the records are less than 4% and sales value will not have major impact with inclusion of these records

```

---
Category and sales comparison
```{r}
df %>% group_by(Category) %>% summarise(N = n(),
                                        Sales_Max = max(Sales),
                                        Sales_min = min(Sales),
                                        Sales_Total = sum(Sales),
                                        Sales_mean = mean(Sales))
                                        
df %>% filter(Sales<10000) %>% group_by(Category) %>% summarise(N = n(),
                                        Sales_Max = max(Sales),
                                        Sales_min = min(Sales),
                                        Sales_Total = sum(Sales),
                                        Sales_mean = mean(Sales))



# drastic change in maximum sales amount, mean has around 8% change
# removing observations with >=10000 sales amount
# only 5 observations will get deleted

df = df %>% filter(Sales < 10000)

# deleting unwanted data sets
product_id = NULL
product_list = NULL
similar_id = NULL

```

---
```{r}
# total countries in the data set
table(df$Country) 
# only united states

# total states in the data set
# visualising top sales contributors
df %>% group_by(State) %>% summarise(Total_Sales = sum(Sales)) %>% 
        ggplot() + geom_bar(aes(reorder(State,-Total_Sales),Total_Sales),stat = "identity")+
        theme_minimal() + 
        theme(axis.text.x = element_text(angle = 80, hjust = 1)) + 
        labs(title = 'Statewise sales',x = 'State') 

# California and New York are the top contributors

View(df %>% group_by(State) %>% summarise(N = n(),
                                Sales_Max = max(Sales),
                                Sales_min = min(Sales),
                                Sales_Total = sum(Sales),
                                Sales_mean = mean(Sales))
                                %>% arrange(desc(Sales_Total)))

View(df %>% group_by(State,Category) %>% summarise(N = n(),
                                          Sales_Max = max(Sales),
                                          Sales_min = min(Sales),
                                          Sales_Total = sum(Sales),
                                          Sales_mean = mean(Sales))
                                        %>% arrange(State,desc(Sales_Total)))

# from the top 5 sales contributors only washington has high sales contribution by furniture category
# all other top 4 sales contributord are having technology as highest sales contributor

# relation between quantity and sales
plot(df$Sales,df$Quantity,xlab = "Sales",ylab = "Quantity",main = "Sales vs Quantity")

## which category has maximum quantity sold
df %>% group_by(Category) %>% summarise(total_quantity = sum(Quantity),
                                        total_sales = sum(Sales)) %>% arrange(desc(total_quantity))

# Office supplies are having high quantity numbers where as technology sales amount is higher
# may be technology items are more expensive

```

---
```{r}
# pulling out year and month out of order date variable
library(lubridate)
summary(df$`Order Date`)
df = df %>% mutate(month_year = format(as.Date(`Order Date`),"%Y-%m"))

```

---
```{r}
# Exploring sub category vaiable
df %>% group_by(`Sub-Category`) %>% summarise(Freq = n(),
                                                   Qty = sum(Quantity)) %>% 
         arrange(desc(Qty))

# Papers and Binders are the top 2 items in terms of quantity sold
df %>% group_by(`Sub-Category`) %>% summarise(Freq = n(),
                                                   Qty = sum(Quantity),
                                                    Total_Sale = sum(Sales)) %>% 
         arrange(desc(Total_Sale))

# sales is dominated by phones and chairs sub category
# checking the sales percentage share of each category

Sales_Share = df %>% group_by(Category) %>% 
              summarise(Sales = sum(Sales))

Sales_Share %>% mutate(Share_Per = paste0(round(Sales/sum(Sales)*100,2),"%")) %>% 
         arrange(desc(Share_Per))

# all three categories seems to have quiet equal share

```

---
```{r}
# Buidling time series model
# forecasting at united states level
ts_data = df %>%group_by(month_year) %>% summarise(Total_Sales = sum(Sales)) %>% 
                arrange(month_year) %>% select(Total_Sales)

# converting data into ts class
ts_df = ts(ts_data,frequency = 12,start = c(2014,1))
end(ts_df)

plot.ts(ts_df) 
plot(decompose(ts_df))
```

---
```{r}
# statistical check for stationarity
adf.test(ts_df) #p-value is 0.07 
```

---
```{r}
# differencing time series
ts_diff = diff(ts_df,differences = 2)
plot.ts(ts_diff)
adf.test(ts_diff)
# time series looks stationary

# selecting p,d,q
acf(ts_diff, lag.max = 20)
pacf(ts_diff, lag.max = 20)

# selecting p as 1 and q as 0
ts_model = arima(ts_diff,order = c(1,2,1))
ts_model

```

---
```{r}
# forecasting for next two years
ts_forecast = forecast:::forecast.Arima(ts_model,h=6)
plot(ts_forecast)
```

---
```{r}
# checking residuals
acf(ts_forecast$residuals, lag.max = 20)
plot.ts(ts_forecast$residuals) # residuals looks random
acf(ts_forecast$residuals)

Box.test(ts_forecast$residuals,lag = 20,type = 'Ljung-Box')
# p value is < 0.05
# residuals are not independent
# forecasting needs improvement
# model is not performing well
```


---
```{r}
# using auto.arima function
ts_auto_arima = auto.arima(ts_df,seasonal = TRUE,stationary = TRUE)
ts_auto_arima

auto_arima_forecast = forecast:::forecast.Arima(ts_auto_arima,h=12)
plot(auto_arima_forecast)
auto_arima_forecast

library(TSPred)
# predicted vd actual graph
plotarimapred(ts_df,auto_arima_forecast$fitted,xlim = c(2014,2017))


# checking residuals auto corelation
acf(auto_arima_forecast$residuals)
# no auto corelation
```

---
```{r}
# checking residuals
plot(auto_arima_forecast$residuals) 
# residuals are random

Box.test(auto_arima_forecast$residuals,lag = 20,type = "Ljung-Box")
# p value is 0.33 i.e. we can not reject null hypothesis
# we can say residuals are independent
# model is performing well
```

---
```{r}
# mean squared error
mean(auto_arima_forecast$residuals)
sqrt(mean(auto_arima_forecast$residuals))

# checking mean absolute percentage error
mean((abs((ts_df-auto_arima_forecast$fitted)/ts_df))*100)
# the model is off by 29%
# the model is performing well at (100-29) 71%

```

---
Forecast for  overall united states
```{r}
forecasted_values = as.data.frame(auto_arima_forecast)
forecasted_values
```

---
Creating forecast at a category level
```{r}
Category_Forecast = function(data,Catg)
{
  ts_data = data %>% filter(Category == Catg) %>%
          group_by(month_year) %>% summarise(Total_Sales = sum(Sales)) %>%
          arrange(month_year) %>% select(Total_Sales)

  # converting data into ts class
  ts_df = ts(ts_data,frequency = 12,start = c(2014,1))
  end(ts_df)

  print(plot.ts(ts_df,main="Time series plot")) 
  #plot(decompose(ts_df))

  # statistical check for stationarity
  adf.test(ts_df) 
  
  # checking with auto arima
  ts_auto_arima = auto.arima(ts_df,seasonal = TRUE,stationary = TRUE)

  auto_arima_forecast = forecast:::forecast.Arima(ts_auto_arima,h=12)
  #print(auto_arima_forecast)
  
  # predicted vd actual graph
  library(TSPred)
  plotarimapred(ts_df,auto_arima_forecast$fitted,xlim = c(2014,2017),
                main = "Actual vs Fprecasted Graph")
   
  # checking residuals auto corelation
  acf(auto_arima_forecast$residuals)
  
  print("checking stationarity of residuals")
  print(adf.test(auto_arima_forecast$residuals))
  
  # checking residuals
  plot(auto_arima_forecast$residuals,main="Residual plot",xlab="Residuals") 
  # residuals are random
  
  print(plot(auto_arima_forecast))
  
  # mean squared error
  mean(auto_arima_forecast$residuals)
  sqrt(mean(auto_arima_forecast$residuals))
  
  # checking mean absolute percentage error
  print("Mean absolute percentage error")
  print(mean((abs((ts_df-auto_arima_forecast$fitted)/ts_df))*100))
  
  print(Box.test(auto_arima_forecast$residuals,lag = 20,type = "Ljung-Box"))

  forecasted_values = as.data.frame(auto_arima_forecast)
  
  return(forecasted_values)
}

Technology_forecast = Category_Forecast(data = df,Catg = "Technology")
Technology_forecast

```

---
```{r}
# forecasting for furniture product
Furniture_forecast = Category_Forecast(data = df,Catg = "Furniture")
Furniture_forecast

```


---
Office supplies forecast
```{r}
OS_forecast = Category_Forecast(df,"Office Supplies")
OS_forecast
```


